---
title: "MWF_HW2"
author: "Savannah McNair"
date: "2025-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1. Get EVS data. The data you want is ZA7500. You can find it here: https://search.gesis.org/research_data/ZA7500 (10%)
I made a research account under my UMD email, smcnair1@umd.edu to download the data. I downloaded the SPSS version of the data, and stored it in the folder I will be working out of for this assignment.

#2. Write an overall report as a pdf.
The two variables of interest are v72 and v80 (I’ll let you discover what exactly they measure). Other variables
you should keep are age, education, sex and country. For the overall report I would like to have a short
intro about the research and data (one paragraph), a section with descriptives that includes max two tables
(e.g., one for continuous variables and one for categorical variables), a section that includes two graphs: how
the two variables of interest change with age, and a section with two regression models explaining the two
dependent variables. The model should include: age, age squared, sex and education. The results from both
models should be presented in a single table. Add a sentence or two interpreting each table and graph.

The European Values Study is a large-scale, cross-national and longitudinal survey research program on how Europeans think about family, work, religion, politics, and society. Repeated every nine years in an increasing number of countries, the survey provides insights into the ideas, beliefs, preferences, attitudes, values, and opinions of citizens all over Europe.

##• clean data for analysis (20%)
necessary packages
```{r}
library(haven)
library(stargazer)
```

pull in data 
```{r}
location <- "C:/Users/smcnair1/Desktop/MWF_HW2/ZA7500_v5-0-0.sav/ZA7500_v5-0-0.sav"
dataset <- read_sav(location)
```

examine data 
```{r}
names(dataset)
```

Observations from first glance at data:
age variable is "age"
education of respondent recoded into 3 categories is "v243_r"
sex of respondent "v225"
country (with abbreviation) is "c_abrv"
v72 is child suffers with working mother (1-4)
v80 is jobs are scarce... giving (nation) priority (1-5)

_DE columns are duplicates with many NAs, I won't need these for my analyses

remove _DE columns
```{r}
dataset <- dataset[, !grepl('_DE', names(dataset))]
```

remove incomplete questionnaires
```{r}
dataset <- dataset[dataset$fmissings != 1, ]
```

keep only the columns of interest (above are the variables I selected)
```{r}
dataset <- dataset[, c("age", "v243_r", "v225", "c_abrv", "v72", "v80")]
```

remove entries where education level question was refused and remove rows with NAs
```{r}
dataset <- dataset[dataset$v243_r != 66, ]
dataset <- na.omit(dataset)
```

rename columns, keep attributes
```{r}
colnames(dataset) <- c("age", "education", "sex", "country", "child_suffers_mom", "job_scarcity")
```

now we have a clean, relevant, analyzable dataset!

##• descriptive tables, graphs and regression tables (20%)
necessary packages
```{r}
library(dplyr)
#install.packages("summarytools")
library(summarytools)
```

descriptive tables
continuous - given that child suffers and job scarcity variables are bipolar lickert scales with numerical values, im making the analytical decision to include them here
```{r}
continuous_summary <- data.frame(
  Variable = c("age", "child_suffers_mom", "job_scarcity"),
  Mean = c(mean(dataset$age), mean(dataset$child_suffers_mom), mean(dataset$job_scarcity)),
  Median = c(median(dataset$age), median(dataset$child_suffers_mom), median(dataset$job_scarcity)),
  SD = c(sd(dataset$age), sd(dataset$child_suffers_mom), sd(dataset$job_scarcity)),
  Min = c(min(dataset$age), min(dataset$child_suffers_mom), min(dataset$job_scarcity)),
  Max = c(max(dataset$age), max(dataset$child_suffers_mom), max(dataset$job_scarcity)),
  Range = c(max(dataset$age) - min(dataset$age), 
            max(dataset$child_suffers_mom) - min(dataset$child_suffers_mom),
            max(dataset$job_scarcity) - min(dataset$job_scarcity))
)
```
```{r}
print(continuous_summary)
```
We can learn a number of things from this table. We know that the mean and median age are very close in this population, hovering around 50 with a standard deviation o 17 years (fairly spread out). We know that over time the a people answer around 2 or 3 out of 4 (agree->disagree) for child_suffers_mom, with lower variability. We know people answer close to 2 out of 5 (agree->disagree) with higher variability on job_scarcity. 

categorical 
create summary
```{r}
get_categorical_summary <- function(variable, var_name) {
  freq_table <- table(variable)  # Frequency table
  prop_table <- prop.table(freq_table) * 100  # Percentage table
  
  summary_df <- data.frame(
    Variable = var_name,
    Level = names(freq_table),
    Frequency = as.integer(freq_table),
    Percentage = round(as.numeric(prop_table), 2)
  )
  
  return(summary_df)
}
```

create summmary for each and append to one table
```{r}
education_summary <- get_categorical_summary(dataset$education, "education")
sex_summary <- get_categorical_summary(dataset$sex, "sex")
country_summary <- get_categorical_summary(dataset$country, "country")

combined_summary <- rbind(education_summary, sex_summary, country_summary)

print(combined_summary)
```

This table tells us a number of things. First, that education is slightly skewed towards higher education (3) over low (1), with most falling in the medium category (2). Less people over all are men (1, 44%) as compared with women (2, 56%). Most countries make up able 2% of the overall sample. The Netherlands (NL) and Denmark (DK) are substantially larger.

graphs
necessary packages
```{r}
library(ggplot2)
```

child suffers when mom works
```{r}
ggplot(dataset, aes(x = age, fill = factor(child_suffers_mom))) +
  geom_bar(position = "fill") +
  labs(x = "Age", y = "Proportion", fill = "Agree (1) -> Disagree (4)") +
  ggtitle("Agreement that 'when a mother works for pay, the children suffers', by Age") +
  scale_y_continuous(labels = scales::percent) +  
  theme_minimal()
```
With age, people tend to agree with the statement "when a mother works for pay, the children suffers" more. The agreement is more moderate, as most people with age will select "agree" (2) over "agree strongly" (1).

job scarcity
```{r}
ggplot(dataset, aes(x = age, fill = factor(job_scarcity))) +
  geom_bar(position = "fill") +
  labs(x = "Age Group", y = "Proportion", fill = "Agree (1) -> Disagree (5)") +
  ggtitle("Agreement that 'when jobs are scarce, employers should give 
  priority to {NATIONALITY} people over immigrants', by Age") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
```
With age, people will tend to agree with "when jobs are scarce, employers should give priority to {NATIONALITY} people over immigrants" more. This agreement is more pronounced, as more people will choose "Agree strongly" over "Agree", and the disagreement is vastly smaller in comparison.

regression tables

create new variables of interest
```{r}
dataset$age_squared <- (dataset$age)^2
```

create models
```{r}
#model 1 child suffers when mom works
model1 <- lm(child_suffers_mom ~ age + age_squared + sex + education, data = dataset)

#model 2 job scarcity impact on immigrants
model2 <- lm(job_scarcity ~ age + age_squared + sex + education, data = dataset)
```

get summaries
```{r}
summary_table <- stargazer(model1, model2, type = "text")
print(summary_table)
```
Model 1 (child_suffers_mom) suggests that age, sex, and education are significantly associated with whether people think a child suffers when the mother works, with sex and education being positively related to the outcome and age showing a slight negative relationship.

Model 2 (job_scarcity) finds that education and sex are significant predictors of peoples thoughts that job scarcity should give priority in jobs to nationals over immigrants. Education has a positive association with this variable, and being male slightly reduces this impact.

However, both models have low R² values, meaning they only explain a small portion of the variance in the dependent variables, and there may be other factors not captured by these models that influence the outcomes.

##• two version of the report, one showing no R code (for policy makers) and one that also shows your code (for statisticians) (10%)

#3. Country level automated report in html.
##• the 33 reports and the syntax that produces them (20%)
##• dynamic (i.e., changes depending on the data) interpretation of results (10%)